---
title: "Custom code: DML and TMLE"
---

{{< include ../_setup.qmd >}}

Cross-fit $\hat\alpha$ from `rieszreg` plus a cross-fit outcome regression $\hat\mu$ are the two ingredients of a one-step / DML / TMLE estimator for any linear-functional estimand. This page builds both estimators on the Lalonde NSW dataset in ~50 lines.

## Data: Lalonde NSW (treated) + CPS (controls)

The Dehejia-Wahba dataset: NSW-treated job-trainees vs CPS observational controls, outcome `re78` (1978 earnings). The standard covariates are age, education, race, marital status, schooling, and pre-period earnings.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from causaldata import nsw_mixtape, cps_mixtape

COVARIATES = ["age", "educ", "black", "hisp", "marr", "nodegree", "re74", "re75"]

treated  = nsw_mixtape.load_pandas().data
treated  = treated[treated["treat"] == 1].copy()
controls = cps_mixtape.load_pandas().data
df = pd.concat([treated, controls], ignore_index=True)
df = df.rename(columns={"treat": "a"})
df["y"] = df["re78"]
df = df[["a", "y"] + COVARIATES].copy()
df["a"] = df["a"].astype(float)
print(f"n = {len(df)}  ({(df['a']==1).sum()} treated, {(df['a']==0).sum()} control)")
```

## R

```{r}
library(reticulate)
cd_py <- import("causaldata")
COVARIATES <- c("age", "educ", "black", "hisp", "marr", "nodegree", "re74", "re75")

treated  <- py_to_r(cd_py$nsw_mixtape$load_pandas()$data)
treated  <- treated[treated$treat == 1, ]
controls <- py_to_r(cd_py$cps_mixtape$load_pandas()$data)
df <- rbind(treated, controls)
df$a <- as.numeric(df$treat)
df$y <- df$re78
df <- df[, c("a", "y", COVARIATES)]
cat(sprintf("n = %d  (%d treated, %d control)\n",
            nrow(df), sum(df$a == 1), sum(df$a == 0)))
```

:::

## Cross-fit nuisances

Both $\hat\mu$ and $\hat\alpha$ go through `cross_val_predict` so every prediction used in the influence function is out-of-fold.

::: {.panel-tabset group="lang"}

## Python

```{python}
import xgboost as xgb
from sklearn.model_selection import KFold, cross_val_predict
from rieszreg import ATE
from rieszboost import RieszBooster

cv = KFold(n_splits=5, shuffle=True, random_state=0)

# Outcome regression μ̂(A, X)
X_full = df[["a"] + COVARIATES].to_numpy(dtype=float)
y      = df["y"].to_numpy(dtype=float)
mu_hat = xgb.XGBRegressor(
    objective="reg:squarederror", learning_rate=0.05,
    max_depth=4, reg_lambda=1.0, n_estimators=300,
    random_state=0, verbosity=0,
)
mu_oof = cross_val_predict(mu_hat, X_full, y, cv=cv)

# Riesz representer α̂(A, X) for ATE
booster = RieszBooster(
    estimand=ATE(treatment="a", covariates=tuple(COVARIATES)),
    n_estimators=2000, early_stopping_rounds=20, validation_fraction=0.2,
    learning_rate=0.05, max_depth=3, reg_lambda=10.0, random_state=0,
)
alpha_hat = cross_val_predict(booster, df[["a"] + COVARIATES], cv=cv)

# Counterfactual μ̂(1, X) and μ̂(0, X) need a model fit on the full data.
mu_full = mu_hat.fit(X_full, y)
X1 = np.column_stack([np.ones(len(df)),  df[COVARIATES].to_numpy(dtype=float)])
X0 = np.column_stack([np.zeros(len(df)), df[COVARIATES].to_numpy(dtype=float)])
mu1 = mu_full.predict(X1)
mu0 = mu_full.predict(X0)
```

## R

```{r}
library(xgboost)
suppressPackageStartupMessages(library(rsample))
set.seed(0)

# Shared 5-fold split — used for both μ̂ and α̂ so they're aligned.
splits <- vfold_cv(df, v = 5)$splits

# Outcome regression μ̂(A, X) via xgboost (R) with 5-fold cross-fit.
X_full <- as.matrix(df[, c("a", COVARIATES)])
y      <- df$y
mu_oof <- numeric(nrow(df))
for (split in splits) {
  te_idx <- complement(split)
  tr_idx <- setdiff(seq_len(nrow(df)), te_idx)
  bst <- xgb.train(
    params = list(objective = "reg:squarederror", eta = 0.05,
                  max_depth = 4, lambda = 1.0, verbosity = 0),
    data = xgb.DMatrix(X_full[tr_idx, , drop = FALSE], label = y[tr_idx]),
    nrounds = 300
  )
  mu_oof[te_idx] <- predict(bst, X_full[te_idx, , drop = FALSE])
}

# Riesz representer α̂ via the same 5-fold loop, calling $fit / $predict on
# the R6 booster per fold.
make_booster <- function() RieszBooster$new(
  estimand = ATE(treatment = "a", covariates = COVARIATES),
  n_estimators = 2000L, early_stopping_rounds = 20L, validation_fraction = 0.2,
  learning_rate = 0.05, max_depth = 3L, reg_lambda = 10.0
)
alpha_hat <- numeric(nrow(df))
for (split in splits) {
  m <- make_booster()
  m$fit(analysis(split))
  alpha_hat[complement(split)] <- m$predict(assessment(split))
}
booster <- make_booster()  # used by the TMLE step below to produce α at counterfactual treatment levels

# Counterfactual μ̂(1, X) and μ̂(0, X) from a full-data fit.
mu_full <- xgb.train(
  params = list(objective = "reg:squarederror", eta = 0.05,
                max_depth = 4, lambda = 1.0, verbosity = 0),
  data = xgb.DMatrix(X_full, label = y),
  nrounds = 300
)
X1 <- cbind(a = 1, as.matrix(df[, COVARIATES]))
X0 <- cbind(a = 0, as.matrix(df[, COVARIATES]))
mu1 <- predict(mu_full, X1)
mu0 <- predict(mu_full, X0)
```

:::

## DML / one-step

The efficient influence function for ATE is
$\mathrm{IF}(O) = m(\hat\mu)(Z) + \hat\alpha(Z)\,(Y - \hat\mu(Z))$,
with $m(\hat\mu)(Z) = \hat\mu(1, X) - \hat\mu(0, X)$. Average it for $\hat\psi$; its sample variance gives the SE.

::: {.panel-tabset group="lang"}

## Python

```{python}
eif = (mu1 - mu0) + alpha_hat * (y - mu_oof)
psi = float(eif.mean())
se  = float(np.sqrt(np.var(eif, ddof=1) / len(eif)))
ci  = (psi - 1.96 * se, psi + 1.96 * se)
print(f"ψ̂_ATE  = {psi:.2f}")
print(f"SE     = {se:.2f}")
print(f"95% CI = [{ci[0]:.2f}, {ci[1]:.2f}]")
```

## R

```{r}
eif <- (mu1 - mu0) + alpha_hat * (y - mu_oof)
psi <- mean(eif)
se  <- sqrt(var(eif) / length(eif))
ci  <- c(psi - 1.96 * se, psi + 1.96 * se)
cat(sprintf("psi_ATE = %.2f\nSE      = %.2f\n95%% CI  = [%.2f, %.2f]\n",
            psi, se, ci[1], ci[2]))
```

:::

The experimental NSW-only ATE benchmark is roughly \$1,794 (Dehejia-Wahba 1999); the NSW-treated vs CPS-controls comparison drifts further from that benchmark when covariates fail to absorb the experimental-vs-observational gap.

## TMLE

TMLE solves the same EIF=0 estimating equation as one-step but does it by *targeting* $\hat\mu$: fit a one-parameter fluctuation $\mu^\star(\cdot) = \hat\mu(\cdot) + \epsilon \hat\alpha(\cdot)$ on a "clever covariate" $\hat\alpha$ to drive the empirical mean of $\hat\alpha(O)(Y - \mu^\star(O))$ to zero, then take the plug-in $\hat\psi = (1/n) \sum \mu^\star(1, X) - \mu^\star(0, X)$. For continuous $Y$ the targeting is a single linear regression of the outcome residual on $\hat\alpha$.

To target the counterfactual predictions we also need $\hat\alpha$ at $A=1$ and $A=0$. Refit the booster on the full sample and predict at the two counterfactual treatment levels.

::: {.panel-tabset group="lang"}

## Python

```{python}
booster.fit(df[["a"] + COVARIATES])
df_a1 = df.assign(a=1.0)
df_a0 = df.assign(a=0.0)
alpha_1 = booster.predict(df_a1[["a"] + COVARIATES])
alpha_0 = booster.predict(df_a0[["a"] + COVARIATES])

# Fluctuation: ε from OLS of (y - μ̂_oof) on α̂_oof, no intercept.
eps = float(np.dot(alpha_hat, y - mu_oof) / np.dot(alpha_hat, alpha_hat))

# Targeted counterfactual regressions and observed-row regression.
mu1_star    = mu1   + eps * alpha_1
mu0_star    = mu0   + eps * alpha_0
mu_obs_star = mu_oof + eps * alpha_hat

psi_tmle = float((mu1_star - mu0_star).mean())
eif_tmle = (mu1_star - mu0_star - psi_tmle) + alpha_hat * (y - mu_obs_star)
se_tmle  = float(np.sqrt(np.var(eif_tmle, ddof=1) / len(eif_tmle)))
ci_tmle  = (psi_tmle - 1.96 * se_tmle, psi_tmle + 1.96 * se_tmle)
print(f"ψ̂_TMLE = {psi_tmle:.2f}")
print(f"SE     = {se_tmle:.2f}")
print(f"95% CI = [{ci_tmle[0]:.2f}, {ci_tmle[1]:.2f}]")
```

## R

```{r}
df_a1 <- df; df_a1$a <- 1
df_a0 <- df; df_a0$a <- 0
booster$fit(df)
alpha_1 <- booster$predict(df_a1)
alpha_0 <- booster$predict(df_a0)

eps          <- sum(alpha_hat * (y - mu_oof)) / sum(alpha_hat^2)
mu1_star     <- mu1   + eps * alpha_1
mu0_star     <- mu0   + eps * alpha_0
mu_obs_star  <- mu_oof + eps * alpha_hat

psi_tmle <- mean(mu1_star - mu0_star)
eif_tmle <- (mu1_star - mu0_star - psi_tmle) + alpha_hat * (y - mu_obs_star)
se_tmle  <- sqrt(var(eif_tmle) / length(eif_tmle))
ci_tmle  <- c(psi_tmle - 1.96 * se_tmle, psi_tmle + 1.96 * se_tmle)
cat(sprintf("psi_TMLE = %.2f\nSE       = %.2f\n95%% CI   = [%.2f, %.2f]\n",
            psi_tmle, se_tmle, ci_tmle[1], ci_tmle[2]))
```

:::

For continuous $Y$, one-step and TMLE point estimates coincide to $O_p(n^{-1})$. TMLE's plug-in is a parameter of an updated regression, which preserves bounds (e.g. probabilities stay in $[0, 1]$ for binary outcomes via the logistic-fluctuation variant — replace the OLS targeting step with a logistic-link single-covariate GLM on the bounded outcome).

## Other estimands

The pattern generalizes verbatim. Swap the estimand factory and the $m(\hat\mu)(Z)$ formula:

| Estimand | $m(\hat\mu)(Z)$ | Riesz factory |
|---|---|---|
| ATT | $A(\hat\mu(1, X) - \hat\mu(0, X))$, then divide by $\hat{P}(A=1)$ | `ATT(...)` |
| TSM at $a^\star$ | $\hat\mu(a^\star, X)$ | `TSM(level=a_star)` |
| Additive shift $\delta$ | $\hat\mu(A + \delta, X) - \hat\mu(A, X)$ | `AdditiveShift(delta)` |

Worked examples for each: [`rieszboost/examples/`](https://github.com/rieszreg/rieszboost/tree/main/examples) (`tsm.py`, `stochastic_intervention.py`, `lee_schuler/binary_dgp.py`).

## See also

- [`rieszboost/examples/lalonde_ate.py`](https://github.com/rieszreg/rieszboost/blob/main/examples/lalonde_ate.py) — runnable Python script this page is based on.
- [`rieszboost/examples/lee_schuler/binary_dgp.py`](https://github.com/rieszreg/rieszboost/blob/main/examples/lee_schuler/binary_dgp.py) — `eee_ate` and `eee_att` helper functions, simulation-based coverage check.
- [Estimation → external packages](external.qmd) — DoubleML / EconML / tlverse routes.
