---
title: "Estimands"
---

{{< include _setup.qmd >}}

A `FiniteEvalEstimand` describes the linear functional you want to fit. It holds three pieces of information: the column names $\alpha$ is indexed by (`feature_keys`), any per-row payload columns $m$ needs such as pre-sampled treatments for stochastic interventions (`extra_keys`), and the `m(alpha)(z)` callable.

`feature_keys` declares the input schema of $\alpha$ — the columns the underlying learner consumes when scoring a row. `extra_keys` declares non-feature columns that $m$ reaches into per row (for example, the list of pre-sampled treatments used by `StochasticIntervention`); they are passed through to $m$ as part of $z$ but are not features of $\alpha$.

This page covers the six built-in factories and how to write your own. The factories live in `rieszreg`; `rieszboost`, `krrr`, and `forestriesz` all re-export them so `from rieszboost import ATE`, `from krrr import ATE`, and `from forestriesz import ATE` resolve to the same object.

<!-- 
I think we should more clearly differentiate between m() operators that reduce to a finite linear combination of point evaluations of \mu and other kinds of linear operators. For the former class things are quite clear and it seems like data augmentation can pretty generically handle things. For estimands that don't fit this pattern, like the stochastic intervention, or an average derivative effect, we can still handle them but with more restrictions on the learner. This needs to be clear in the docs but also in the code, i.e. there needs to be a subclass of estimand that represents the finite linear combination of point eval case and the tracer only needs to handle that subclass (I think). 

There may be another subclass to handle integrals, and maybe another for derivatives would make sense. The integrals can be handled with sampling in which case they sort of get replaced by / transformed to a finite evals case with more data? Derivatives may also come down to a finite evals case after a transformation that gets handled by the learner (eg autodiff). Need help thinking through this (are there other classes? What else is a linear operator besides finite linear combination, integration, and differentiation?) and how the Estimands and code should be structured to make this clear and to enforce the right things.

Whatever conclusion/structure we come to, it will have to be documented clearly.
-->

## Finite-evaluation estimands

While the Riesz representer theorem applies to any continuous linear functional, the estimands that are easiest to support computationally are those that reduce to a finite linear combination of point evaluations of $\mu$. For example, the ATE is $\mathbb{E}[\mu(1, X) - \mu(0, X)]$, which reduces to the two point evaluations $\mu(1, x)$ and $\mu(0, x)$ for each row's covariates $x$. Many of our built-in estimands fit this pattern. `rieszreg` can work with other estimands as well, but since they are more rarely of interest and come at a computational cost or are only supported by some backends, we first focus on the finite-evaluation class.

We say an estimand $\psi = E[m(\mu)(Z,Y)]$ is of finite-evaluation type if there exist fixed functions $b: \mathbb R^k \times \mathcal Z \times \mathcal Y \to \mathbb R$ and $a: (\mathcal Z \times \mathcal Y) \to \mathcal Z^k$ such that

$$
m(\mu)(z, y) = b(\vec\mu(a(z,y)), z, y)
$$

Where $\vec\mu$ denotes elementwise application of $\mu$ to its vector-valued input. The function $a$ selects $k$ points at which to evaluate $\mu$, based on the values of $z$ and $y$. For example, for the ATE, $k=2$ and $a(z,y) = ((1, z_x), (0, z_x))$ selects the treated and control potential-covariate pairs. The function $b$ then forms the contrast $b(\mu_1, \mu_0, z, y) = \mu_1 - \mu_0$. 

The first observation we can make is that if an estimand is of this type (and $m$ is a linear operator, which we already must assume), it can be written as 

$$
m(\mu)(z, y) = c(z,y)^\top \vec\mu(a(z,y))
$$

for some function $c: (\mathcal Z \times \mathcal Y) \to \mathbb R^k$ by the linearity of $m$. In other words, $m$ must be a finite *linear combination* of point evaluations of $\mu$. In the ATE example, $c(z,y) = (1, -1)$ gives the contrast weights for the treated and control evaluations of $\mu$.

Therefore for these estimands we can rewrite the empirical Bregman-Riesz loss. Define the $k$ pseudo-datapoints $Z_{ij} = a_j(Z_i, Y_i)$ with coefficients $C_{ij} = c_j(Z_i, Y_i)$ for each row $i$. Then:

$$
\begin{aligned}
\mathcal L_{h,n}(\alpha)
&= \sum_{i=1}^n \left[\tilde h(\alpha(Z_i)) - m(h' \circ \alpha)(Z_i,Y_i) \right] \\
&= \sum_{i=1}^n \left[\tilde h(\alpha(Z_i)) - \sum_{j=1}^k C_{ij} \cdot h'(\alpha(Z_{ij}))  \right]
\end{aligned}
$$

The loss depends on $\alpha$ only through its values at $\{Z_i\} \cup \{Z_{ij}\}$, at most $n(k+1)$ points. The coefficients $C_{ij}$ depend on data but not on $\alpha$ or on the potential function for the loss, so they can be computed once as a preprocessing step. The advantage of this is that, during fitting, evaluating the empirical loss only requires computing the values of $\alpha$ at these points. Methods that depend on a gradient of the empricial loss with respect to either some parameters of $\alpha$ (like `riesznet`) or $\alpha$ itself (`rieszboost`) inherit this, as do methods that depend on an analytic solution to the empirical loss minimization problem (`krrr`).

### The augmented dataset

The easiest way to keep track of the loss, deriviatves, etc. during training is to create an augmented dataset consisting of the points $\{Z_i\} \cup \{Z_{ij}\}$, along with the appropriate coefficients $C_{ij}$.

Some pseudo-datapoints may coincide with the original observation. For ATE, $Z_{i1} = (1, x_i)$ and $Z_{i2} = (0, x_i)$. A treated unit ($a_i = 1$) has $Z_{i1} = Z_i$, so two of the three evaluation points are the same. Let $D_{ij} = \mathbf{1}[Z_{ij} = Z_i]$ indicate overlap. We can split the $h'$ sum into overlapping and non-overlapping terms:

$$
\mathcal L_{h,n}(\alpha) = \sum_{i=1}^n \left[\tilde h(\alpha(Z_i)) - \sum_{j:\, D_{ij}=1} C_{ij}\, h'(\alpha(Z_i))\right] - \sum_{i=1}^n \sum_{j:\, D_{ij}=0} C_{ij}\, h'(\alpha(Z_{ij}))
$$

At each distinct evaluation point, the loss combines a $\tilde h$-weight and an $h'$-weight into a single term. `build_augmented` collects these distinct points for each row and stores their coefficients. Let $r \in 1 \dots |\{Z_i\} \cup \{Z_{ij}\}|$ index an enumeration of the augmented data points. Each of the augmented data points $Z_r$ corresponds to either a $Z_{ij}$ or a $Z_i$, either of which corresponds to some original index row $i$. Write $i_r$ to index the origin row corresponding to augmented row $r$. The two weights for the augmented row are:

$$
D_r = \mathbf{1}[Z_r = Z_{i_r}],
\qquad
C_r = -\sum_{j \,:\, Z_{i_r, j} = Z_r} C_{i_r, j}.
$$

The $\tilde h$-weight $D_r$ is $1$ exactly when $Z_r$ is the original observation, and the $h'$-weight $C_r$ collects $-C_{ij}$ from every pseudo-datapoint of row $i_r$ that lands at $Z_r$ — an empty sum (giving $C_r = 0$) when no pseudo-datapoint coincides with $Z_{i_r}$. The per-row contribution to the empirical loss is then $L_r = D_r\, \tilde h(\alpha(Z_r)) + C_r\, h'(\alpha(Z_r))$, and $\mathcal L_{h,n}(\alpha) = \sum_r L_r$.

For ATE with an untreated unit ($a_i = 0$): before merging there are three evaluation points — $Z_i = (0, x_i)$ and pseudo-datapoints $(1, x_i)$, $(0, x_i)$. The pseudo-datapoint $(0, x_i)$ coincides with $Z_i$ ($D_{i2} = 1$), so they merge into one entry with $\tilde h$-weight 1 and $h'$-weight $-C_{i2} = +1$. The other pseudo-datapoint $(1, x_i)$ stays separate with $\tilde h$-weight 0 and $h'$-weight $-C_{i1} = -1$. The result is two augmented rows instead of three. A treated unit merges the other pseudo-datapoint instead, also producing two rows.

It can also happen that none of the pseudo-datapoints coincide with $Z_i$ — the union $\{Z_i\} \cup \{Z_{ij}\}$ then has $k+1$ distinct entries and no merging occurs. For example, `TSM(level=1)` on an untreated row ($a_i = 0$) has $Z_i = (0, x_i)$ and a single pseudo-datapoint $(1, x_i)$; they are distinct, so $Z_i$ enters the augmented dataset as its own row with $\tilde h$-weight $1$ and $h'$-weight $0$ (no pseudo-datapoint lands there), and $(1, x_i)$ enters as a separate row with $\tilde h$-weight $0$ and $h'$-weight $-C_{i1} = -1$. The general rule is the same in every case: include $Z_i$ once, then add each pseudo-datapoint $Z_{ij}$ — merging into an existing entry when $Z_{ij}$ matches a point already in the dataset for row $i$, otherwise as a new row.

To make this concrete, take the ATE estimand on $n=3$ rows:

| $i$ | $A_i$ | $X_i$ |
|----:|------:|------:|
| 1   | 0     | 0.2   |
| 2   | 1     | 0.5   |
| 3   | 1     | 0.8   |

For each row, the three evaluation points $\{Z_i,\, Z_{i1} = (1, X_i),\, Z_{i2} = (0, X_i)\}$ collapse to two distinct entries after merging. The augmented dataset has $2n = 6$ rows; each row $r$ carries the evaluation point $(A_r, X_r)$, the $\tilde h$-weight $D_r$, the $h'$-weight $C_r$, and contributes a term $L_r = D_r \cdot \tilde h(\alpha(A_r, X_r)) + C_r \cdot h'(\alpha(A_r, X_r))$ to the empirical loss:

| from $i$ | $A_r$ | $X_r$ | $D_r$ | $C_r$ | $L_r$ |
|---------:|------:|------:|------:|------:|:------|
| 1        | 0     | 0.2   | $1$   | $+1$  | $\tilde h(\alpha(0, 0.2)) + h'(\alpha(0, 0.2))$ |
| 1        | 1     | 0.2   | $0$   | $-1$  | $-h'(\alpha(1, 0.2))$ |
| 2        | 1     | 0.5   | $1$   | $-1$  | $\tilde h(\alpha(1, 0.5)) - h'(\alpha(1, 0.5))$ |
| 2        | 0     | 0.5   | $0$   | $+1$  | $+h'(\alpha(0, 0.5))$ |
| 3        | 1     | 0.8   | $1$   | $-1$  | $\tilde h(\alpha(1, 0.8)) - h'(\alpha(1, 0.8))$ |
| 3        | 0     | 0.8   | $0$   | $+1$  | $+h'(\alpha(0, 0.8))$ |

Rows $i=2$ and $i=3$ are treated, so the merging happens on the $(1, X_i)$ pseudo-datapoint instead of $(0, X_i)$, flipping which row carries the $\tilde h$-weight. Summing the $L_r$ column recovers the empirical loss:

$$
\mathcal L_{h,n}(\alpha) = \sum_{r=1}^{6} L_r = \sum_{i=1}^{3} \big[\tilde h(\alpha(Z_i)) - h'(\alpha(1, X_i)) + h'(\alpha(0, X_i))\big].
$$

`build_augmented` produces this table directly: it stores the two coefficients as `is_original` (the $\tilde h$-weight $D_r$) and `potential_deriv_coef` (the $h'$-weight $C_r$) on the returned `AugmentedDataset`.

```{python}
import numpy as np
import pandas as pd
from rieszreg import ATE, build_augmented

df = pd.DataFrame({"a": [0, 1, 1], "x": [0.2, 0.5, 0.8]})
df
```

```{python}
estimand = ATE(treatment="a", covariates=("x",))
aug = build_augmented(df.to_dict("records"), estimand)

aug_df = pd.DataFrame({
    "i": aug.origin_index + 1,
    "A": aug.features[:, 0],
    "X": aug.features[:, 1],
    "D": aug.is_original,
    "C": aug.potential_deriv_coef,
})
aug_df
```

Pick a candidate $\alpha(a, x) = a + x$ and the squared loss ($\tilde h(t) = t^2$, $h'(t) = 2t$). The per-row contribution becomes $L_r = D \cdot \alpha^2 + 2 C \cdot \alpha$:

```{python}
F = aug_df["A"] + aug_df["X"]                    # α evaluated at each augmented point
aug_df["L_r"] = aug.is_original * F**2 + 2.0 * aug.potential_deriv_coef * F
aug_df
```

```{python}
aug_df["L_r"].sum()                              # empirical loss
```

The same number drops out of the original row-wise form $\sum_i [\alpha(Z_i)^2 - 2\alpha(1, X_i) + 2\alpha(0, X_i)]$:

```{python}
(df["a"] + df["x"])**2 - 2*(1 + df["x"]) + 2*(0 + df["x"])
```

```{python}
((df["a"] + df["x"])**2 - 2*(1 + df["x"]) + 2*(0 + df["x"])).sum()
```

### Built-in Estimands

| Estimand | $m(\alpha)(z)$ | Notes |
|---|---|---|
| `ATE(treatment, covariates)` | $\alpha(1, x) - \alpha(0, x)$ | Average treatment effect — see [`rieszboost/examples/lee_schuler/binary_dgp.py`](https://github.com/rieszreg/rieszreg/blob/main/rieszboost/examples/lee_schuler/binary_dgp.py) |
| `ATT(treatment, covariates)` | $a \cdot (\alpha(1, x) - \alpha(0, x))$ | ATT *partial-estimand* surface. Full ATT divides by $P(A=1)$ — combine with a delta-method EIF downstream. |
| `TSM(level, treatment, covariates)` | $\alpha(\text{level}, x)$ | Treatment-specific mean — see [`rieszboost/examples/tsm.py`](https://github.com/rieszreg/rieszreg/blob/main/rieszboost/examples/tsm.py) |
| `AdditiveShift(delta, ...)` | $\alpha(a + \delta, x) - \alpha(a, x)$ | Continuous-treatment shift effect — see [`rieszboost/examples/lee_schuler/continuous_dgp.py`](https://github.com/rieszreg/rieszreg/blob/main/rieszboost/examples/lee_schuler/continuous_dgp.py) |
| `LocalShift(delta, threshold, ...)` | $\mathbf{1}\{a < \text{threshold}\} \cdot (\alpha(a + \delta, x) - \alpha(a, x))$ | LASE *partial-estimand* surface; same caveat as ATT. |

The R and Python factories take the same arguments.

::: {.panel-tabset group="lang"}

## Python

```{python}
from rieszreg import (
    ATE, ATT, TSM, AdditiveShift, LocalShift, StochasticIntervention,
)
ATE(treatment="a", covariates=("x",)).name
TSM(level=1, treatment="a", covariates=("x",)).name
LocalShift(delta=0.3, threshold=0.7).name
```

## R

```{r}
ATE(treatment = "a", covariates = "x")
TSM(level = 1, treatment = "a", covariates = "x")
LocalShift(delta = 0.3, threshold = 0.7)
```

:::

## Other Estimands



`StochasticIntervention` is a `FiniteEvalEstimand`: once each row's $K$ samples are fixed, the operator is the finite linear combination $m(\alpha)(z) = (1/K) \sum_k \alpha(a'_k, x)$. Pre-draw the samples once before fitting, and the estimand fits like any other.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from rieszboost import RieszBooster
from rieszreg import StochasticIntervention

rng = np.random.default_rng(0)
n = 1000
x = rng.uniform(0, 1, n)
a = rng.uniform(0, 1, n)
df = pd.DataFrame({"a": a, "x": x})

# Modified-treatment policy: a' ~ Normal(a + 0.1, 0.2), 20 samples per row
df["shift_samples"] = [
    rng.normal(a_i + 0.1, 0.2, size=20) for a_i in df["a"]
]

booster = RieszBooster(
    estimand=StochasticIntervention(samples_key="shift_samples"),
    n_estimators=200, learning_rate=0.05, max_depth=4,
).fit(df)

booster.predict(df).mean()
```

## R

```{r}
set.seed(0)
n  <- 1000
x  <- runif(n); a <- runif(n)
df <- data.frame(a = a, x = x)

# list-column with 20 samples per row from Normal(a + 0.1, 0.2)
df$shift_samples <- lapply(df$a, function(a_i) rnorm(20, a_i + 0.1, 0.2))

booster <- RieszBooster$new(
  estimand = StochasticIntervention(samples_key = "shift_samples"),
  n_estimators = 200L, learning_rate = 0.05, max_depth = 4L
)
booster$fit(df)
mean(booster$predict(df))
```

:::

`StochasticIntervention` sets `extra_keys=("shift_samples",)` so the estimator pulls that column through to `m` along with the features.

## Custom estimands

To fit a finite-evaluation functional not covered by the built-ins, write the operator `m(alpha)(z)` and wrap it in a `FiniteEvalEstimand`. The tracer extracts the linear-form structure automatically.

`m` must be a finite linear combination of point evaluations of $\alpha$, built with `+`, `-`, scalar `*`, and calls to `alpha(...)`. Squaring, adding constants, integrals, and other non-linear operations raise.

::: {.callout-note title="Python-only"}
Custom `m()` is Python-only. From R, define `m` in Python via `reticulate::py_run_string` and pass the resulting `FiniteEvalEstimand` to your estimator. See the [R interface page](r-interface.qmd) for the recipe.
:::

```{python}
from rieszreg import FiniteEvalEstimand
from rieszboost import RieszBooster
import numpy as np, pandas as pd

# A weighted contrast: 2x the ATE at level a=2 minus the level at a=1.
def m_weighted_contrast(alpha):
    def inner(z):
        return 2.0 * (alpha(a=2, x=z["x"]) - alpha(a=0, x=z["x"])) \
               - alpha(a=1, x=z["x"])
    return inner

est = FiniteEvalEstimand(
    feature_keys=("a", "x"),
    m=m_weighted_contrast,
    name="WeightedContrast",
)
print(est)
```

Pass the estimand to any backend's estimator class:

```{python}
rng = np.random.default_rng(0)
df = pd.DataFrame({
    "a": rng.choice([0, 1, 2], size=500).astype(float),
    "x": rng.uniform(0, 1, 500),
})

booster = RieszBooster(estimand=est, n_estimators=100, max_depth=3).fit(df)
booster.predict(df)[:5]
```

### What the tracer rejects

The errors point at the offending operation:

```{python}
#| error: true
from rieszreg import FiniteEvalEstimand, trace

def bad_squared(alpha):
    return lambda z: alpha(a=1, x=z["x"]) ** 2  # nonlinear in alpha

trace(FiniteEvalEstimand(feature_keys=("a", "x"), m=bad_squared), {"a": 0, "x": 0.5})
```

```{python}
#| error: true
def bad_constant(alpha):
    return lambda z: alpha(a=1, x=z["x"]) + 1.0  # constant offset

trace(FiniteEvalEstimand(feature_keys=("a", "x"), m=bad_constant), {"a": 0, "x": 0.5})
```

For estimands with integrals or derivatives of $\alpha$ (outside the finite-linear-form algebra), Monte-Carlo-sample and use `StochasticIntervention` if applicable.

## Feature keys, extra keys

The `FiniteEvalEstimand` dataclass has these fields:

```python
@dataclass
class FiniteEvalEstimand(Estimand):
    feature_keys: tuple[str, ...]      # columns alpha is indexed by
    m: Callable[..., Any]              # the linear functional, m(alpha)(z)
    extra_keys: tuple[str, ...] = ()   # per-row payload columns (e.g. shift_samples)
    name: str = "custom"
    factory_spec: dict | None = None   # set by built-in factories so the estimand
                                       # round-trips through save / load
```

The estimator reads `feature_keys` and `extra_keys` to assemble the row dicts passed to `m`. `feature_keys` is the input schema of $\alpha$ (the underlying learner needs a fixed feature set); `extra_keys` lets `m` reach payload columns that are not part of $\alpha$'s input space.

## Initialization

Every `RieszEstimator` subclass takes an `init=` constructor arg that controls the α-space starting value of the model. The orchestrator converts `init` to an η-space `base_score` via `loss.alpha_to_eta`; backends use `base_score` as a prediction offset.

The default — `init=None` — sets α to the **empirical loss-minimizing constant**. For any Bregman loss with strictly convex $h$, the constant $a^*$ that minimizes the per-row Riesz loss
$$
P_n L(a) \;=\; \tilde h(a) \;-\; h'(a) \cdot \bar m, \qquad \bar m = P_n[m(\alpha = 1)(Z, Y)]
$$
satisfies the FOC $\tilde h'(a) = h''(a) \bar m$. Using $\tilde h(t) = t h'(t) - h(t)$ we get $\tilde h'(t) = t h''(t)$, so the FOC collapses to $a^* = \bar m$. The result is **loss-independent** — the only loss-specific step is projecting $\bar m$ into the loss's α-domain when its codomain is restricted. Each `Loss` exposes `best_constant_init(m_bar)` for that projection.

For the built-in estimands $\bar m$ is structurally constant (doesn't depend on the data realization):

| Estimand | $m(\alpha = 1)(Z, Y)$ per row | $\bar m$ |
|---|---|---|
| `ATE`, `ATT`, `AdditiveShift`, `LocalShift` | $1 + (-1) = 0$ | $0$ |
| `TSM(v)` | $1$ | $1$ |
| `StochasticIntervention` | $K \cdot (1/K) = 1$ | $1$ |

For custom estimands $\bar m$ can depend on the data; the orchestrator computes it from the training rows after the validation split (no peeking) by tracing each row and averaging the sum of trace coefficients.

Override the default by passing a float:

```python
booster = RieszBooster(estimand=ATE(), init=0.0)        # explicit zero baseline
booster = RieszBooster(estimand=TSM(level=1), init=2.5) # informed prior
```

::: {.callout-note title="Boundary cases"}
For `BernoulliLoss` + `TSM` (or any density-ratio estimand with $\bar m = 1$) the loss-minimizing constant sits on the upper boundary, and the projection clips to $1 - \varepsilon$. The sigmoid is saturated at that init and gradients are tiny, so iterative training may stall. Pass `init=0.5` (or another interior value) when this matters. The `KLLoss` / `BoundedSquaredLoss` boundaries can hit the same issue when $\bar m$ lands exactly on a bound.
:::