---
title: "RieszReg"
title-block-style: none
engine: knitr
---

{{< include _setup.qmd >}}

![](banner.svg){.hero-banner fig-alt="RieszReg — Riesz regression for semiparametric inference"}

`rieszreg` estimates the Riesz representer $\alpha$ of a linear estimand $\psi = \mathbb{E}[m(\mu)(Z,Y)]$ — the building block of one-step, TMLE, and DML inference — with gradient boosting ([`rieszboost`](backends/boosting.qmd)), kernel ridge ([`krrr`](backends/kernel.qmd)), random forests ([`forestriesz`](backends/forest.qmd)), or neural nets ([`riesznet`](backends/neural.qmd)). Think of it as the `sklearn`, `tidymodels` or `superlearner` for Riesz regression.

`rieszreg` ships built-in factories for ATE, ATT, TSM, additive shifts, local shifts, stochastic interventions, and arbitrary user-defined estimands; the squared Riesz loss plus the wider Bregman family (KL, Bernoulli, bounded squared); and sklearn-native tuning, model selection, and cross-fitting. Python and R, with bitwise-identical predictions on the same input.

## What is Riesz regression?

You want to estimate some causal or structural estimand — say, the average treatment effect $\psi = \mathbb{E}[\mu(1, X) - \mu(0, X)]$, where $\mu(a, x) = \mathbb{E}[Y \mid A = a, X = x]$. A regression $\hat\mu$ for $\mu$ is biased; the standard fix uses a weight. For ATE, the weight is the inverse-propensity score:

$$
\alpha(a, x) \;=\; \frac{a}{\pi(x)} - \frac{1-a}{1-\pi(x)},
$$

Under the [Riesz representation theorem](intro.qmd), similar weights exist for many different estimands. `rieszreg` learns these weights directly from data: no propensity model, no truncation, etc. The same machinery handles continuous shifts, stochastic interventions, and many custom estimands a user can write down — for which closed-form weights could otherwise involve tricky or slow density estimates. These weights are then suitable for downstream use in weighting estimators or 

For the longer derivation and a survey of the closed-form weights for each built-in estimand, see [What is Riesz regression?](intro.qmd). 

## Quickstart

Fit a Riesz representer for ATE on a synthetic binary-treatment dataset, select between two backends with cross-validation, and use the cross-fit $\hat\alpha$ for a DML point estimate and Wald CI.

::: {.panel-tabset group="lang"}

## Python

```{python}
#| code-fold: true
#| code-summary: "Simulate"
import os
os.environ.update({"OMP_NUM_THREADS": "1", "MKL_NUM_THREADS": "1"})
import numpy as np
import pandas as pd

rng = np.random.default_rng(0)
n = 2000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02 * x - x**2 + 4 * np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi)
mu = 5 * x + 9 * x * a + 5 * np.sin(x * np.pi) + 25 * (a - 2)
y = mu + rng.normal(0, 1.0, n)
df = pd.DataFrame({"a": a.astype(float), "x": x, "y": y})
true_alpha = a / pi - (1 - a) / (1 - pi)
```

We start with a data frame `df` with treatment `a`, covariate `x`, outcome `y`. In this case, since this is simulated data, we know the true Riesz representer $\alpha_0$ to evaluate against later.

```{python}
import torch
torch.set_num_threads(1)
from rieszreg import ATE, BoundedSquaredLoss
from rieszboost import RieszBooster
from riesznet import RieszNet

estimand = ATE(treatment="a", covariates=("x",))
loss     = BoundedSquaredLoss(lo=-50.0, hi=50.0)

boost = RieszBooster(
    estimand=estimand, loss=loss,
    n_estimators=200, learning_rate=0.05, max_depth=3,
)
net   = RieszNet(
    estimand=estimand, loss=loss,
    hidden_sizes=(32, 32), epochs=60, learning_rate=1e-2,
)
```

Wrap the candidates in a `Pipeline` and let `GridSearchCV` pick the winner via inner CV; nest that selector inside `cross_val_predict` so each outer fold runs its own inner model selection on the training portion only. The result `alpha_hat` is out-of-fold — every row was predicted by a model fit without seeing it.

```{python}
from sklearn.model_selection import GridSearchCV, KFold, cross_val_predict
from sklearn.pipeline import Pipeline

cv = KFold(n_splits=5, shuffle=True, random_state=0)
selector = GridSearchCV(
    Pipeline([("riesz", boost)]),
    param_grid=[{"riesz": [boost, net]}],
    cv=3, refit=True,
)
alpha_hat = cross_val_predict(selector, df, cv=cv)
```

### Sanity check

```{python}
#| echo: false
#| fig-cap: "Out-of-fold $\\hat\\alpha$ vs the closed-form $\\alpha_0$ on the synthetic ATE DGP."
import matplotlib.pyplot as plt
fig, ax = plt.subplots()
ax.scatter(true_alpha, alpha_hat, s=4, alpha=0.4)
lim = max(abs(true_alpha).max(), abs(alpha_hat).max())
ax.plot([-lim, lim], [-lim, lim], "k--", lw=1)
ax.set_xlabel(r"$\alpha_0$ (closed-form)")
ax.set_ylabel(r"$\hat\alpha$ (cross-fit)")
plt.show()
```

### DML point estimate and CI

```{python}
#| code-fold: true
#| code-summary: "Cross-fit outcome regression"
from sklearn.ensemble import GradientBoostingRegressor

mu_model = GradientBoostingRegressor(
    n_estimators=300, learning_rate=0.05, max_depth=3, random_state=0,
)
X_full = df[["a", "x"]].to_numpy()
mu_oof = cross_val_predict(mu_model, X_full, df["y"].to_numpy(), cv=cv)

# Counterfactual predictions need a model fit on the full data.
mu_full = mu_model.fit(X_full, df["y"].to_numpy())
mu1 = mu_full.predict(np.column_stack([np.ones(n), x]))
mu0 = mu_full.predict(np.column_stack([np.zeros(n), x]))
```

```{python}
eif  = (mu1 - mu0) + alpha_hat * (df["y"].to_numpy() - mu_oof)
psi  = float(eif.mean())
se   = float(np.sqrt(np.var(eif, ddof=1) / n))
ci   = (psi - 1.96 * se, psi + 1.96 * se)
print(f"ψ̂_ATE  = {psi:.3f}")
print(f"SE     = {se:.3f}")
print(f"95% CI = [{ci[0]:.3f}, {ci[1]:.3f}]")
```

## R

```{r}
#| code-fold: true
#| code-summary: "Load riesznet R wrapper"
.project_root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = getwd())
suppressMessages(pkgload::load_all(
  file.path(.project_root, "../riesznet/r/riesznet"), quiet = TRUE))
```

```{r}
#| code-fold: true
#| code-summary: "Simulate"
set.seed(0)
n     <- 2000
x     <- runif(n)
prop  <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a     <- rbinom(n, 1, prop)
mu_dgp <- 5 * x + 9 * x * a + 5 * sin(x * pi) + 25 * (a - 2)
y     <- mu_dgp + rnorm(n, 0, 1)
df    <- data.frame(a = as.numeric(a), x = x, y = y)
true_alpha <- a / prop - (1 - a) / (1 - prop)
```

```{r}
estimand <- ATE(treatment = "a", covariates = "x")
loss     <- BoundedSquaredLoss(lo = -50, hi = 50)
make_boost <- function() RieszBooster$new(
  estimand = estimand, loss = loss,
  n_estimators = 200L, learning_rate = 0.05, max_depth = 3L
)
make_net <- function() RieszNet$new(
  estimand = estimand, loss = loss,
  hidden_sizes = c(32L, 32L), epochs = 60L, learning_rate = 0.01
)
```

Use [`rsample`](https://rsample.tidymodels.org/) (tidymodels) for fold splits, and a small loop for the nested cross-fit + per-fold model selection. Each outer fold runs its own inner-CV winner pick on its training portion only.

```{r}
suppressPackageStartupMessages(library(rsample))

cv_outer <- vfold_cv(df, v = 5)
alpha_hat <- numeric(nrow(df))
for (split in cv_outer$splits) {
  tr <- analysis(split); te <- assessment(split); te_idx <- complement(split)
  cv_inner <- vfold_cv(tr, v = 3)
  inner_score <- function(make) mean(vapply(cv_inner$splits, function(s) {
    m <- make(); m$fit(analysis(s)); m$score(assessment(s))
  }, numeric(1)))
  best_make <- if (inner_score(make_boost) >= inner_score(make_net)) make_boost else make_net
  best <- best_make(); best$fit(tr)
  alpha_hat[te_idx] <- best$predict(te)
}
```

Sanity check $\hat\alpha$ against the closed-form $\alpha_0$:

```{r}
#| echo: false
#| fig-cap: "Out-of-fold $\\hat\\alpha$ vs the closed-form $\\alpha_0$ (R)."
lim <- max(abs(c(true_alpha, alpha_hat)))
plot(true_alpha, alpha_hat, pch = 16, cex = 0.4, col = adjustcolor("black", 0.4),
     xlim = c(-lim, lim), ylim = c(-lim, lim),
     xlab = expression(alpha[0]~"(closed-form)"),
     ylab = expression(hat(alpha)~"(cross-fit)"))
abline(0, 1, lty = 2)
```

Cross-fit outcome regression $\hat\mu$ via xgboost, then assemble the DML estimate and Wald CI:

```{r}
#| code-fold: true
#| code-summary: "Cross-fit outcome regression"
suppressPackageStartupMessages(library(xgboost))

xgb_params <- list(objective = "reg:squarederror", eta = 0.05,
                   max_depth = 3, lambda = 1, verbosity = 0)
mu_oof <- numeric(nrow(df))
for (split in cv_outer$splits) {
  tr <- analysis(split); te <- assessment(split); te_idx <- complement(split)
  bst <- xgb.train(
    params = xgb_params,
    data = xgb.DMatrix(as.matrix(tr[, c("a", "x")]), label = tr$y),
    nrounds = 300
  )
  mu_oof[te_idx] <- predict(bst, as.matrix(te[, c("a", "x")]))
}
mu_full <- xgb.train(
  params = xgb_params,
  data = xgb.DMatrix(as.matrix(df[, c("a", "x")]), label = df$y),
  nrounds = 300
)
mu1 <- predict(mu_full, cbind(a = 1, as.matrix(df[, "x", drop = FALSE])))
mu0 <- predict(mu_full, cbind(a = 0, as.matrix(df[, "x", drop = FALSE])))
```

```{r}
eif <- (mu1 - mu0) + alpha_hat * (df$y - mu_oof)
psi <- mean(eif)
se  <- sqrt(var(eif) / length(eif))
ci  <- c(psi - 1.96 * se, psi + 1.96 * se)
cat(sprintf("psi_ATE = %.3f\nSE      = %.3f\n95%% CI  = [%.3f, %.3f]\n",
            psi, se, ci[1], ci[2]))
```

:::

For TMLE on top of these nuisances, and integration with `DoubleML` and other downstream packages, see [Estimation](estimation/index.qmd). For backend-specific tuning, see [Backends](backends/index.qmd).

::: {.callout-tip title="Code execution"}
Every code block on this page ran at build time. The numbers and figures came from the version of `rieszreg`, `rieszboost`, `krrr`, `forestriesz`, and `riesznet` that produced this page.
:::
