---
title: "Losses"
---

{{< include _setup.qmd >}}

The Bregman-Riesz loss generalizes the squared Riesz loss via a strictly convex potential $\phi$. `rieszreg` ships four built-in `LossSpec` classes; backends consume per-row gradient and Hessian sourced from the loss spec.

::: {.callout-note title="Backend support"}
- **Boosting** ([`rieszboost`](backends/boosting.qmd)): all four losses.
- **Kernel ridge** ([`krrr`](backends/kernel.qmd)): `SquaredLoss` only вҖ” non-quadratic losses need a Newton iteration on the kernel system.
- **Random forest, augmentation-style** (`forestriesz.AugForestRieszRegressor`): all four losses. Tree structure from the squared MSE criterion, then a per-leaf Newton replaces each leaf's stored Оё with the Bregman optimum.
- **Random forest, moment-style** (`forestriesz.ForestRieszRegressor`): `SquaredLoss` only вҖ” the moment-style per-leaf gradient is not exposed by the loss API.
- **Neural network** ([`riesznet`](backends/neural.qmd)): all four losses. Each loss's per-row Bregman form is implemented in autograd-friendly torch ops; gradient parity vs the analytic loss is checked at 1e-8.

Pass an unsupported loss and the backend raises at fit time with a clear error.
:::

## Bregman-Riesz Losses
<!-- claude, replace hines et al and kato and chernozhukov below with actual links to papers -->
Hines et al. [year] show that for a convex, differentiable function $h: \mathbb R \to \mathbb R$ (called $F$ in their paper), the minimizer of a population loss based on $h$ (the average Bregman divergence between $\alpha$ and $\alpha_0$ with potential $h$) can be expressed in a form that does not involve the true Riesz representer $\alpha_0$. 
$$
\begin{aligned}
\arg\min_\alpha \mathcal L_g(\alpha) 
&= \arg\min_\alpha \mathbb E[h(\alpha_0(Z)) - h(\alpha(Z)) - h'(\alpha(Z))(\alpha_0)(Z) - \alpha(Z))] \\
&= \arg\min_\alpha \mathbb E[\tilde h(\alpha(Z)) - m(h' \circ \alpha)(Z,Y)] \\
\end{aligned}
$$

where $\tilde h(t) = t h'(t) - h(t)$. Kato {year} and Hines [year] both give some intuition for why this loss makes sense and what sensible choices of $h$ are in different cases. When $h(t) = t^2$ this reduces to the standard Riesz loss of Chernozhukov et al [year]. `rieszreg` implements several Bregman-Riesz losses and also allows users to define their own by defining the function $h$.

<!-- add what you can distill from this to your notation skill -->

## Loss table

| Loss | Potential $\phi$ | Link $\eta \to \alpha$ | Domain | Use for |
|---|---|---|---|---|
| `SquaredLoss` | $t^2$ | identity | $\mathbb{R}$ | Default. |
| `KLLoss` | $t \log t$ | exp | $(0, \infty)$ | Density-ratio estimands (TSM, IPSI). Guarantees $\hat\alpha > 0$. Requires non-negative $m$-coefficients. |
| `BernoulliLoss` | $t \log t + (1-t)\log(1-t)$ | sigmoid | $(0, 1)$ | $\alpha_0$ known to lie in $(0, 1)$. Same coefficient-sign restriction as `KLLoss`. |
| `BoundedSquaredLoss(lo, hi)` | $t^2$ + sigmoid-scaled link | $\mathrm{lo} + (\mathrm{hi}-\mathrm{lo}) \cdot \sigma$ | $(\mathrm{lo}, \mathrm{hi})$ | Squared-loss fit with prior bounds (e.g. trimmed-propensity ratios). |

`SquaredLoss` is the standard Lee-Schuler / Chernozhukov objective. Per-row loss in $\alpha$-space: $a \cdot \alpha^2 + b \cdot \alpha$.

## `KLLoss` for density-ratio estimands

`KLLoss` ($\phi = t \log t$, exp link) is matched to density-ratio estimands. The exponential link keeps $\hat\alpha > 0$.
<!-- I don't understand what we gain from the link, can't it always be composed with h()? So shouldn't we just define the composition directly instead of leaving more params for the user to try to make sense of? the link doesn't add any generality, right? -->

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from rieszboost import RieszBooster
from rieszreg import KLLoss, TSM

rng = np.random.default_rng(0)
n = 1000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(8 * x - 4)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})

booster = RieszBooster(
    estimand=TSM(level=1),
    loss=KLLoss(max_eta=50),
    n_estimators=300,
    learning_rate=0.05,
    max_depth=3,
).fit(df)

alpha_hat = booster.predict(df)
print(f"min ОұМӮ = {alpha_hat.min():.4f}")
print(f"mean ОұМӮ on treated rows = {alpha_hat[a == 1].mean():.3f}")
```

## R

```{r}
set.seed(0)
n  <- 1000
x  <- runif(n)
pi <- 1 / (1 + exp(-(8 * x - 4)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

booster <- RieszBooster$new(
  estimand = TSM(level = 1),
  loss = KLLoss(max_eta = 50),
  n_estimators = 300L,
  learning_rate = 0.05,
  max_depth = 3L
)
booster$fit(df)
alpha_hat <- booster$predict(df)
cat(sprintf("min alpha_hat = %.4f\n", min(alpha_hat)))
cat(sprintf("mean alpha_hat on treated rows = %.3f\n",
            mean(alpha_hat[a == 1])))
```

:::

`KLLoss` requires all augmented `b` coefficients $\le 0$, equivalent to all $m$-coefficients being non-negative. The check happens at fit time and raises with a clear error if violated.

<!-- I don't think we want to be talking about coefficients yet, right now we haven't restricted the estimands yet so that this makes sense at all -->

## `BoundedSquaredLoss` for hard prior bounds

If you know $\alpha_0 \in [\text{lo}, \text{hi}]$ (e.g. you've trimmed propensities to $[\varepsilon, 1-\varepsilon]$ so $1/\pi \in [1, 1/\varepsilon]$), `BoundedSquaredLoss(lo, hi)` keeps predictions in that box via a sigmoid-scaled link. The per-row loss in $\alpha$-space matches `SquaredLoss`; the link reparameterizes $\eta$ so $\hat\alpha$ saturates at the bounds rather than extrapolating past them.

::: {.panel-tabset group="lang"}

## Python

```{python}
from rieszreg import BoundedSquaredLoss
booster = RieszBooster(
    estimand=TSM(level=1),
    loss=BoundedSquaredLoss(lo=0.0, hi=12.0),
    n_estimators=200, learning_rate=0.05, max_depth=3,
).fit(df)
alpha_hat = booster.predict(df)
print(f"alpha_hat in ({alpha_hat.min():.3f}, {alpha_hat.max():.3f})")
```

## R

```{r}
booster <- RieszBooster$new(
  estimand = TSM(level = 1),
  loss = BoundedSquaredLoss(lo = 0.0, hi = 12.0),
  n_estimators = 200L, learning_rate = 0.05, max_depth = 3L
)
booster$fit(df)
alpha_hat <- booster$predict(df)
cat(sprintf("alpha_hat in (%.3f, %.3f)\n",
            min(alpha_hat), max(alpha_hat)))
```

:::

::: {.callout-tip title="Pick bounds tightly"}
Generous bounds make the sigmoid link saturate over most of $\eta$, so boosting steps barely move predictions. Set `lo` and `hi` to your prior on $\alpha_0$.
:::

`BernoulliLoss` is the same construction with $\text{lo}=0$, $\text{hi}=1$ and the Bernoulli Bregman potential. Use it when $\alpha_0$ is a probability.

## Initialization

Every `RieszEstimator` subclass takes an `init=` constructor arg that controls the Оұ-space starting value of the model. The orchestrator converts `init` to an О·-space `base_score` via `loss.alpha_to_eta`; backends use `base_score` as a prediction offset (boosting starts the additive sum from there, neural backends add it inside the forward pass, kernel-ridge folds it into the linear-system right-hand side, forests add it to the leaf prediction).

The default вҖ” `init=None` вҖ” sets Оұ to the **empirical loss-minimizing constant**.

For any Bregman loss with strictly convex $\phi$, the constant $a^*$ that minimizes the per-row Riesz loss
$$
P_n L(a) \;=\; \psi(a) \;-\; \varphi'(a) \cdot \bar m, \qquad \bar m = P_n[m(Z, 1)]
$$
satisfies the FOC $\psi'(a) = \varphi''(a) \bar m$. Using $\psi(t) = t \varphi'(t) - \varphi(t)$ we get $\psi'(t) = t \varphi''(t)$, so the FOC collapses to $a^* = \bar m$. The result is **loss-independent** вҖ” the only loss-specific step is projecting $\bar m$ into the loss's Оұ-domain when its codomain is restricted (KL: $\alpha > 0$; Bernoulli: $\alpha \in (0,1)$; BoundedSquared: $\alpha \in (\mathrm{lo}, \mathrm{hi})$). Each `LossSpec` exposes `best_constant_init(m_bar)` for that projection.

For the built-in estimands $\bar m$ is structurally constant (doesn't depend on the data realization):

| Estimand | $m(z, 1)$ per row | $\bar m$ |
|---|---|---|
| `ATE`, `ATT`, `AdditiveShift`, `LocalShift` | $1 + (-1) = 0$ | $0$ |
| `TSM(v)` | $1$ | $1$ |
| `StochasticIntervention` | $K \cdot (1/K) = 1$ | $1$ |

For custom estimands $\bar m$ can depend on the data; the orchestrator computes it from the training rows after the validation split (no peeking) by tracing each row and averaging the sum of trace coefficients.

Override the default by passing a float:

```python
booster = RieszBooster(estimand=ATE(), init=0.0)        # explicit zero baseline
booster = RieszBooster(estimand=TSM(level=1), init=2.5) # informed prior
```

::: {.callout-note title="Boundary cases"}
For `BernoulliLoss` + `TSM` (or any density-ratio estimand with $\bar m = 1$) the loss-minimizing constant sits on the upper boundary, and the projection clips to $1 - \varepsilon$. The sigmoid is saturated at that init and gradients are tiny, so iterative training may stall. Pass `init=0.5` (or another interior value) when this matters. The `KLLoss` / `BoundedSquaredLoss` boundaries can hit the same issue when $\bar m$ lands exactly on a bound.
:::

<!-- I think we need to move this section to its own page or in estimands page-->

## Custom losses

::: {.callout-note title="Python-only"}
Custom `LossSpec` subclasses are Python-only. From R, define the subclass in Python via `reticulate::py_run_string` and pass the instance via `loss=`.
:::

Implement the `LossSpec` protocol from `rieszreg.losses`:

```python
class LossSpec(Protocol):
    name: str

    def link_to_alpha(self, eta: np.ndarray) -> np.ndarray:
        """Inverse link: convert backend output О· to Оұ."""
    def alpha_to_eta(self, alpha):
        """Forward link: convert Оұ to О· (used for `init=` translation)."""
    def loss_row(self, a, b, alpha):
        """Per-row loss in Оұ space: aВ·ПҲ(Оұ) + (b/2)В·ПҶ'(Оұ)."""
    def gradient(self, a, b, eta):
        """вҲӮloss_row/вҲӮО·, using Оұ = link_to_alpha(О·)."""
    def hessian(self, a, b, eta, hessian_floor):
        """вҲӮВІloss_row/вҲӮО·ВІ (floored)."""
    def best_constant_init(self, m_bar: float) -> float:
        """Loss-minimizing constant Оұ* given mМ„ = E[m(Z, 1)]."""
    def validate_coefficients(self, b: np.ndarray) -> None:
        """Raise if (a, b) violate this loss's domain."""
```

<!-- 
1.can you give an example of what this would mean? "implement this protocol" is a bit abstract for users who aren't used to thinking in terms of protocols. maybe a simple example of a custom loss that implements this protocol, and then how to pass it to a learner?
2. this seems like way too many methods to have to implement. In principle, the loss just depends on a function h: R->R (link is redundant unless you can convince me otherwise). I'd like custom losses to work with just an h() function. Derivatives are needed always but if not defined by the user the should be computable with autodiff. Same for hessian, which I assume we need for some methods. 
3. There should be *nothing* in the loss about "coefficients" or anything like that. Those only apply when we start talking about esitmands that depend only on a finite number of evaluations of alpha, and even then they are just a property of the estimand, not the loss. The loss should just be a function of alpha and the user shouldn't have to worry about how it interacts with the estimand. Those "coefficients" should be computed once the loss is combined with an estimand of the appropraite type. I hope this also clarifies the difference between the data augmentation and moment backends which seems like it might be conceptually misplaced to me
-->

[`SquaredLoss`](https://github.com/rieszreg/rieszreg/blob/main/rieszreg/python/rieszreg/losses/squared.py) and [`KLLoss`](https://github.com/rieszreg/rieszreg/blob/main/rieszreg/python/rieszreg/losses/kl.py) are short reference implementations.

The Bregman extension follows Hines & Miles ([2510.16127](https://arxiv.org/abs/2510.16127)) and Kato ([2601.07752](https://arxiv.org/abs/2601.07752)).
