---
title: "Losses"
---

{{< include _setup.qmd >}}

The Bregman-Riesz loss generalizes the squared Riesz loss via a strictly convex potential $\phi$. `rieszreg` ships four built-in `LossSpec` classes; backends consume per-row gradient and Hessian sourced from the loss spec.

::: {.callout-note title="Backend support"}
The boosting backend ([`rieszboost`](backends/boosting.qmd)) supports all four losses. The kernel ridge backend ([`krrr`](backends/kernel.qmd)) and the random forest backend ([`forestriesz`](backends/forest.qmd)) currently support `SquaredLoss` only — non-quadratic losses require a Newton iteration on the per-leaf or per-system solve. Pass an unsupported loss and the backend raises at fit time with a clear error.
:::

## Loss table

| Loss | Potential $\phi$ | Link $\eta \to \alpha$ | Domain | Use for |
|---|---|---|---|---|
| `SquaredLoss` | $t^2$ | identity | $\mathbb{R}$ | Default. |
| `KLLoss` | $t \log t$ | exp | $(0, \infty)$ | Density-ratio targets (TSM, IPSI). Guarantees $\hat\alpha > 0$. Requires non-negative $m$-coefficients. |
| `BernoulliLoss` | $t \log t + (1-t)\log(1-t)$ | sigmoid | $(0, 1)$ | $\alpha_0$ known to lie in $(0, 1)$. Same coefficient-sign restriction as `KLLoss`. |
| `BoundedSquaredLoss(lo, hi)` | $t^2$ + sigmoid-scaled link | $\mathrm{lo} + (\mathrm{hi}-\mathrm{lo}) \cdot \sigma$ | $(\mathrm{lo}, \mathrm{hi})$ | Squared-loss fit with prior bounds (e.g. trimmed-propensity ratios). |

`SquaredLoss` is the standard Lee-Schuler / Chernozhukov objective. Per-row loss in $\alpha$-space: $a \cdot \alpha^2 + b \cdot \alpha$.

## `KLLoss` for density-ratio targets

`KLLoss` ($\phi = t \log t$, exp link) targets density ratios. The exponential link keeps $\hat\alpha > 0$.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from rieszboost import RieszBooster
from rieszreg import KLLoss, TSM

rng = np.random.default_rng(0)
n = 1000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(8 * x - 4)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})

booster = RieszBooster(
    estimand=TSM(level=1),
    loss=KLLoss(max_eta=50),
    n_estimators=300,
    learning_rate=0.05,
    max_depth=3,
).fit(df)

alpha_hat = booster.predict(df)
print(f"min α̂ = {alpha_hat.min():.4f}")
print(f"mean α̂ on treated rows = {alpha_hat[a == 1].mean():.3f}")
```

## R

```{r}
set.seed(0)
n  <- 1000
x  <- runif(n)
pi <- 1 / (1 + exp(-(8 * x - 4)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

booster <- RieszBooster$new(
  estimand = TSM(level = 1),
  loss = KLLoss(max_eta = 50),
  n_estimators = 300L,
  learning_rate = 0.05,
  max_depth = 3L
)
booster$fit(df)
alpha_hat <- booster$predict(df)
cat(sprintf("min alpha_hat = %.4f\n", min(alpha_hat)))
cat(sprintf("mean alpha_hat on treated rows = %.3f\n",
            mean(alpha_hat[a == 1])))
```

:::

`KLLoss` requires all augmented `b` coefficients $\le 0$, equivalent to all $m$-coefficients being non-negative. The check happens at fit time and raises with a clear error if violated.

::: {.callout-warning title="When KLLoss can be unstable"}
Pure-counterfactual rows ($a=0$, $b<0$) have no minimum in $\eta$ under KL — leaf weights at low-overlap points can extrapolate. Use `SquaredLoss` unless the positivity guarantee is needed.
:::

## `BoundedSquaredLoss` for hard prior bounds

If you know $\alpha_0 \in [\text{lo}, \text{hi}]$ (e.g. you've trimmed propensities to $[\varepsilon, 1-\varepsilon]$ so $1/\pi \in [1, 1/\varepsilon]$), `BoundedSquaredLoss(lo, hi)` keeps predictions in that box via a sigmoid-scaled link. The per-row loss in $\alpha$-space matches `SquaredLoss`; the link reparameterizes $\eta$ so $\hat\alpha$ saturates at the bounds rather than extrapolating past them.

::: {.panel-tabset group="lang"}

## Python

```{python}
from rieszreg import BoundedSquaredLoss
booster = RieszBooster(
    estimand=TSM(level=1),
    loss=BoundedSquaredLoss(lo=0.0, hi=12.0),
    n_estimators=200, learning_rate=0.05, max_depth=3,
).fit(df)
alpha_hat = booster.predict(df)
print(f"alpha_hat in ({alpha_hat.min():.3f}, {alpha_hat.max():.3f})")
```

## R

```{r}
booster <- RieszBooster$new(
  estimand = TSM(level = 1),
  loss = BoundedSquaredLoss(lo = 0.0, hi = 12.0),
  n_estimators = 200L, learning_rate = 0.05, max_depth = 3L
)
booster$fit(df)
alpha_hat <- booster$predict(df)
cat(sprintf("alpha_hat in (%.3f, %.3f)\n",
            min(alpha_hat), max(alpha_hat)))
```

:::

::: {.callout-tip title="Pick bounds tightly"}
Generous bounds make the sigmoid link saturate over most of $\eta$, so boosting steps barely move predictions. Set `lo` and `hi` to your prior on $\alpha_0$.
:::

`BernoulliLoss` is the same construction with $\text{lo}=0$, $\text{hi}=1$ and the Bernoulli Bregman potential. Use it when $\alpha_0$ is a probability.

## Custom losses

::: {.callout-note title="Python-only"}
Custom `LossSpec` subclasses are Python-only. From R, define the subclass in Python via `reticulate::py_run_string` and pass the instance via `loss=`.
:::

Implement the `LossSpec` protocol from `rieszreg.losses`:

```python
class LossSpec(Protocol):
    name: str

    def link_to_alpha(self, eta: np.ndarray) -> np.ndarray:
        """Inverse link: convert boosted output η to α."""
    def alpha_to_eta(self, alpha):
        """Forward link: convert α to η (used for `init=` translation)."""
    def loss_row(self, a, b, alpha):
        """Per-row loss in α space: a·ψ(α) + (b/2)·φ'(α)."""
    def gradient(self, a, b, eta):
        """∂loss_row/∂η, using α = link_to_alpha(η)."""
    def hessian(self, a, b, eta, hessian_floor):
        """∂²loss_row/∂η² (floored)."""
    def default_init_alpha(self) -> float:
        """α-space default for `init=` if user doesn't override."""
    def validate_coefficients(self, b: np.ndarray) -> None:
        """Raise if (a, b) violate this loss's domain."""
```

[`SquaredLoss`](https://github.com/rieszreg/rieszreg/blob/main/rieszreg/python/rieszreg/losses/squared.py) and [`KLLoss`](https://github.com/rieszreg/rieszreg/blob/main/rieszreg/python/rieszreg/losses/kl.py) are short reference implementations.

The Bregman extension follows Hines & Miles ([2510.16127](https://arxiv.org/abs/2510.16127)) and Kato ([2601.07752](https://arxiv.org/abs/2601.07752)).
