---
title: "Tuning and Cross-Fitting"
---

{{< include _setup.qmd >}}

`RieszBooster`, `KernelRieszRegressor`, `ForestRieszRegressor`, `AugForestRieszRegressor`, and `RieszNet` are all subclasses of `rieszreg.RieszEstimator`, which inherits from `sklearn.base.BaseEstimator`. Every sklearn tool that works on an estimator works here: `cross_val_predict`, `GridSearchCV`, `Pipeline`, `clone`, `set_params`, `RandomizedSearchCV`, `HalvingGridSearchCV`, and the rest of `sklearn.model_selection`. Hyperparameter grids, cross-fitting folds, multi-loss comparison, and multi-learner selection all use the same API.

This page uses `RieszBooster` for the single-learner sections and adds `RieszNet` and `ForestRieszRegressor` to the ensemble-selection example. The same patterns work for `KernelRieszRegressor` and `AugForestRieszRegressor` with backend-specific hyperparameters.

## Cross-fitting

Plug-in $\hat\alpha$ in a one-step / TMLE / DML estimator must be out-of-fold. Use `cross_val_predict`:

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from sklearn.model_selection import cross_val_predict
from rieszboost import RieszBooster
from rieszreg import ATE

rng = np.random.default_rng(0)
n = 2000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02 * x - x**2 + 4 * np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})

booster = RieszBooster(
    estimand=ATE(),
    n_estimators=2000,
    early_stopping_rounds=20,
    validation_fraction=0.2,
    learning_rate=0.05,
    max_depth=4,
)

alpha_oof = cross_val_predict(booster, df, cv=5)
alpha_oof.shape, alpha_oof[:5]
```

## R

```{r}
set.seed(0)
n  <- 2000
x  <- runif(n)
pi <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

# Call sklearn directly via reticulate, passing the underlying Python
# estimator from `booster$py`.
sk_ms <- reticulate::import("sklearn.model_selection", convert = FALSE)
booster <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 2000L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2,
  learning_rate = 0.05,
  max_depth = 4L
)
alpha_oof <- as.numeric(reticulate::py_to_r(
  sk_ms$cross_val_predict(booster$py, reticulate::r_to_py(df), cv = 5L)
))
length(alpha_oof); head(alpha_oof)
```

:::

Each fold runs its own internal early-stopping split.

## Hyperparameter search

`.score(X)` returns the negative canonical squared Riesz loss — a fixed yardstick independent of the loss the estimator was trained with, in the spirit of R² for regressors. `GridSearchCV` selects the configuration with the highest score (lowest squared loss). To use a different yardstick, pass `scoring=riesz_scorer(loss=...)`.

::: {.panel-tabset group="lang"}

## Python

```{python}
from sklearn.model_selection import GridSearchCV

grid = GridSearchCV(
    RieszBooster(
        estimand=ATE(),
        n_estimators=300,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    param_grid={"learning_rate": [0.03, 0.05, 0.1], "max_depth": [3, 4, 5]},
    cv=3,
    n_jobs=1,
).fit(df)

print("best params:", grid.best_params_)
print(f"best score (neg squared Riesz loss): {grid.best_score_:.4f}")

best = grid.best_estimator_
alpha_hat = best.predict(df)
alpha_hat[:5]
```

## R

```{r}
sk_ms <- reticulate::import("sklearn.model_selection", convert = FALSE)
base  <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 300L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2
)
grid <- sk_ms$GridSearchCV(
  base$py,
  param_grid = reticulate::dict(
    learning_rate = c(0.03, 0.05, 0.1),
    max_depth = list(3L, 4L, 5L)
  ),
  cv = 3L,
  n_jobs = 1L
)
grid$fit(reticulate::r_to_py(df))
cat("best params:\n"); print(reticulate::py_to_r(grid$best_params_))
cat(sprintf("best score (neg squared Riesz loss): %.4f\n",
            reticulate::py_to_r(grid$best_score_)))
```

:::

## Comparing learners with different losses

Two estimators trained with different losses produce α̂ in the same units (a per-row real value), so the squared yardstick scores them on the same scale. `cross_val_score` results are directly comparable.

```{python}
from sklearn.model_selection import cross_val_score
from rieszreg import BoundedSquaredLoss, SquaredLoss, riesz_scorer

booster_bd = RieszBooster(
    estimand=ATE(), loss=BoundedSquaredLoss(lo=-50.0, hi=50.0),
    n_estimators=300, learning_rate=0.05, max_depth=3,
)
booster_sq = RieszBooster(
    estimand=ATE(), loss=SquaredLoss(),
    n_estimators=300, learning_rate=0.05, max_depth=3,
)

s_bd = cross_val_score(booster_bd, df, cv=3)
s_sq = cross_val_score(booster_sq, df, cv=3)
print(f"BoundedSquared-trained: mean score = {s_bd.mean():.4f}")
print(f"squared-trained:        mean score = {s_sq.mean():.4f}")
```

A single `GridSearchCV` can sweep across losses too. Each candidate is fit with its own loss; all candidates are ranked on the squared yardstick.

```{python}
grid = GridSearchCV(
    RieszBooster(
        estimand=ATE(),
        n_estimators=300, learning_rate=0.05, max_depth=3,
    ),
    param_grid={"loss": [BoundedSquaredLoss(lo=-50.0, hi=50.0), SquaredLoss()]},
    cv=3,
    n_jobs=1,
).fit(df)

print("best loss:", grid.best_params_["loss"])
print(f"best score: {grid.best_score_:.4f}")
```

To rank with a different yardstick, build it with `riesz_scorer`:

```{python}
scorer = riesz_scorer(loss=BoundedSquaredLoss(lo=-50.0, hi=50.0))
s_bd_yard = cross_val_score(booster_bd, df, cv=3, scoring=scorer)
s_sq_yard = cross_val_score(booster_sq, df, cv=3, scoring=scorer)
print(f"BoundedSquared-trained on bounded yardstick: {s_bd_yard.mean():.4f}")
print(f"squared-trained on bounded yardstick:        {s_sq_yard.mean():.4f}")
```

The yardstick must accept the α the estimator predicts. The squared yardstick has unrestricted α-domain, so it works for any learner; the others have restrictions:

| Yardstick | α-domain | Accepts α from learners trained with |
|---|---|---|
| `SquaredLoss` | ℝ | any loss (the default) |
| `KLLoss` | (0, ∞) | KL-trained learners |
| `BernoulliLoss` | (0, 1) | Bernoulli-trained learners |
| `BoundedSquaredLoss(lo, hi)` | (lo, hi) | matching-bounded learners |

Pick a yardstick whose domain accepts every candidate's α. `SquaredLoss` is always safe.

## Cross-fit ensemble selection

Going further than two losses on one backend: the pattern below selects among five learner configurations — two boosting depths, two neural-net architectures, and a forest — using inner 3-fold CV inside each of five outer cross-fitting folds. The inner CV runs entirely on the training portion of each outer fold, so the model selection criterion never sees the held-out data that `cross_val_predict` predicts.

This uses the `Pipeline` idiom for cross-estimator-type selection: wrap the estimator in a single-step `Pipeline` named `"riesz"`, then pass a list of fully configured instances as the value for that key. `GridSearchCV` calls `pipe.set_params(riesz=candidate)`, fits, and scores each one on the canonical squared yardstick.

```{python}
from sklearn.pipeline import Pipeline
from riesznet import RieszNet
from forestriesz import ForestRieszRegressor

estimand = ATE()

# Five fully configured candidates across three learner families.
candidates = [
    # Boosting — depth 3, early stopping
    RieszBooster(
        estimand=estimand,
        n_estimators=2000,
        max_depth=3,
        early_stopping_rounds=20,
        validation_fraction=0.2,
        learning_rate=0.05,
    ),
    # Boosting — depth 5, early stopping
    RieszBooster(
        estimand=estimand,
        n_estimators=2000,
        max_depth=5,
        early_stopping_rounds=20,
        validation_fraction=0.2,
        learning_rate=0.05,
    ),
    # Neural net — small MLP with early stopping
    RieszNet(
        estimand=estimand,
        hidden_sizes=(64, 64),
        epochs=500,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    # Neural net — larger MLP with early stopping
    RieszNet(
        estimand=estimand,
        hidden_sizes=(128, 64),
        epochs=500,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    # Forest Riesz
    ForestRieszRegressor(
        estimand=estimand,
        n_estimators=500,
        min_samples_leaf=10,
        random_state=0,
    ),
]

# Single-step Pipeline so GridSearchCV can swap in any candidate as the step.
pipe = Pipeline([("riesz", candidates[0])])

# Inner CV: 3 folds, selects the best learner+config from the five candidates.
cv_inner = GridSearchCV(
    pipe,
    param_grid=[{"riesz": [c]} for c in candidates],
    cv=3,
    n_jobs=1,
    refit=True,
)

# Outer CV: 5 cross-fitting folds. Each fold:
#   1. Fits cv_inner on 4/5 of the data (inner CV happens here, on that 4/5).
#   2. Predicts the held-out 1/5 using the winner from step 1.
# The inner model selection never touches the held-out slice.
alpha_oof = cross_val_predict(cv_inner, df, cv=5)
print(f"out-of-fold alpha shape: {alpha_oof.shape}")
print(f"first 5 values:          {alpha_oof[:5]}")
```

The call to `cross_val_predict` is identical to the single-learner case. The only addition is the `Pipeline` wrapper and the list-valued `param_grid` entry — everything else is standard sklearn.

The same nested cross-fit + per-fold model selection in R uses [`rsample`](https://rsample.tidymodels.org/) (tidymodels) for the fold splits and a small loop for the inner-CV winner pick. Each outer fold runs its own inner-CV on its training portion only.

```{r}
#| code-fold: true
#| code-summary: "Load forestriesz and riesznet R wrappers"
.project_root <- Sys.getenv("QUARTO_PROJECT_DIR", unset = getwd())
suppressMessages(pkgload::load_all(
  file.path(.project_root, "../forestriesz/r/forestriesz"), quiet = TRUE))
suppressMessages(pkgload::load_all(
  file.path(.project_root, "../riesznet/r/riesznet"),       quiet = TRUE))
```

```{r}
suppressPackageStartupMessages(library(rsample))

estimand <- ATE()
make_boost_d3 <- function() RieszBooster$new(
  estimand = estimand, n_estimators = 2000L, max_depth = 3L,
  early_stopping_rounds = 20L, validation_fraction = 0.2, learning_rate = 0.05
)
make_boost_d5 <- function() RieszBooster$new(
  estimand = estimand, n_estimators = 2000L, max_depth = 5L,
  early_stopping_rounds = 20L, validation_fraction = 0.2, learning_rate = 0.05
)
make_net_64  <- function() RieszNet$new(
  estimand = estimand, hidden_sizes = c(64L, 64L), epochs = 500L,
  early_stopping_rounds = 20L, validation_fraction = 0.2
)
make_net_128 <- function() RieszNet$new(
  estimand = estimand, hidden_sizes = c(128L, 64L), epochs = 500L,
  early_stopping_rounds = 20L, validation_fraction = 0.2
)
make_forest  <- function() ForestRieszRegressor$new(
  estimand = estimand, n_estimators = 500L, min_samples_leaf = 10L,
  random_state = 0L
)
candidates <- list(make_boost_d3, make_boost_d5, make_net_64, make_net_128, make_forest)

cv_outer <- vfold_cv(df, v = 5)
alpha_oof <- numeric(nrow(df))
for (split in cv_outer$splits) {
  tr <- analysis(split); te <- assessment(split); te_idx <- complement(split)
  cv_inner <- vfold_cv(tr, v = 3)
  scores <- vapply(candidates, function(make) {
    mean(vapply(cv_inner$splits, function(s) {
      m <- make(); m$fit(analysis(s)); m$score(assessment(s))
    }, numeric(1)))
  }, numeric(1))
  best_make <- candidates[[which.max(scores)]]
  best <- best_make(); best$fit(tr)
  alpha_oof[te_idx] <- best$predict(te)
}
length(alpha_oof); head(alpha_oof, 5)
```

`ForestRieszRegressor` requires loading `forestriesz` and `riesznet` is needed for the neural candidates; load those R packages alongside `rieszreg` and `rieszboost` per the [install](install.qmd) instructions.

## Other sklearn integrations

Every Riesz estimator exposes the standard `BaseEstimator` interface:

- `clone(estimator)` returns a fresh unfitted copy with the same hyperparameters.
- `Pipeline` wraps a feature-engineering step in front of the estimator.
- `HalvingGridSearchCV`, `RandomizedSearchCV`, `BayesSearchCV` (skopt) work without modification.
- `get_params()` / `set_params()` are used by the search classes to introspect and mutate the configuration.

## See also

- [`rieszboost/examples/lee_schuler/binary_dgp.py`](https://github.com/rieszreg/rieszreg/blob/main/rieszboost/examples/lee_schuler/binary_dgp.py) builds an EEE estimator using `cross_val_predict` to produce out-of-fold $\hat\alpha$.
- [Boosting](backends/boosting.qmd), [kernel](backends/kernel.qmd), [forest](backends/forest.qmd), and [neural](backends/neural.qmd) backend pages cover backend-specific tuning grids.
