---
title: "Tuning and Cross-Fitting"
---

{{< include _setup.qmd >}}

`RieszBooster`, `KernelRieszRegressor`, `ForestRieszRegressor`, `AugForestRieszRegressor`, and `RieszNet` are all subclasses of `rieszreg.RieszEstimator`, which itself inherits from `sklearn.base.BaseEstimator`. Every sklearn tool that works on an estimator works here: `cross_val_predict`, `GridSearchCV`, `Pipeline`, `clone`, `set_params`, `RandomizedSearchCV`, `HalvingGridSearchCV`, and the rest of `sklearn.model_selection`. Hyperparameter grids, cross-fitting folds, and multi-learner selection all use the same API you already know.

This page uses `RieszBooster` and `ForestRieszRegressor`; the same patterns work for `KernelRieszRegressor`, `AugForestRieszRegressor`, and `RieszNet`.

## Cross-fitting

Plug-in $\hat\alpha$ in a one-step / TMLE / DML estimator must be out-of-fold. Use `cross_val_predict`:

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from sklearn.model_selection import cross_val_predict
from rieszboost import RieszBooster
from rieszreg import ATE

rng = np.random.default_rng(0)
n = 2000
x = rng.uniform(0, 1, n)
pi = 1 / (1 + np.exp(-(-0.02 * x - x**2 + 4 * np.log(x + 0.3) + 1.5)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})

booster = RieszBooster(
    estimand=ATE(),
    n_estimators=2000,
    early_stopping_rounds=20,
    validation_fraction=0.2,
    learning_rate=0.05,
    max_depth=4,
)

alpha_oof = cross_val_predict(booster, df, cv=5)
alpha_oof.shape, alpha_oof[:5]
```

## R

```{r}
set.seed(0)
n  <- 2000
x  <- runif(n)
pi <- 1 / (1 + exp(-(-0.02 * x - x^2 + 4 * log(x + 0.3) + 1.5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)

# Call sklearn directly via reticulate, passing the underlying Python
# estimator from `booster$py`.
sk_ms <- reticulate::import("sklearn.model_selection", convert = FALSE)
booster <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 2000L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2,
  learning_rate = 0.05,
  max_depth = 4L
)
alpha_oof <- as.numeric(reticulate::py_to_r(
  sk_ms$cross_val_predict(booster$py, reticulate::r_to_py(df), cv = 5L)
))
length(alpha_oof); head(alpha_oof)
```

:::

Each fold runs its own internal early-stopping split.

## Hyperparameter search

`.score(X)` returns negative held-out Riesz loss, so `GridSearchCV` selects the configuration with the lowest loss.

::: {.panel-tabset group="lang"}

## Python

```{python}
from sklearn.model_selection import GridSearchCV

grid = GridSearchCV(
    RieszBooster(
        estimand=ATE(),
        n_estimators=300,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    param_grid={"learning_rate": [0.03, 0.05, 0.1], "max_depth": [3, 4, 5]},
    cv=3,
    n_jobs=1,
).fit(df)

print("best params:", grid.best_params_)
print(f"best score (neg Riesz loss): {grid.best_score_:.4f}")

best = grid.best_estimator_
alpha_hat = best.predict(df)
alpha_hat[:5]
```

## R

```{r}
sk_ms <- reticulate::import("sklearn.model_selection", convert = FALSE)
base  <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 300L,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2
)
grid <- sk_ms$GridSearchCV(
  base$py,
  param_grid = reticulate::dict(
    learning_rate = c(0.03, 0.05, 0.1),
    max_depth = list(3L, 4L, 5L)
  ),
  cv = 3L,
  n_jobs = 1L
)
grid$fit(reticulate::r_to_py(df))
cat("best params:\n"); print(reticulate::py_to_r(grid$best_params_))
cat(sprintf("best score (neg Riesz loss): %.4f\n",
            reticulate::py_to_r(grid$best_score_)))
```

:::

## Cross-fit ensemble selection

The pattern below selects among five learner configurations — two boosting depths, two neural-network architectures, and a forest — using inner 3-fold CV inside each of five outer cross-fitting folds. The inner CV runs entirely on the training portion of each outer fold, so the model selection criterion never sees the held-out data that `cross_val_predict` predicts.

This uses the `Pipeline` idiom for cross-estimator-type selection: wrap the estimator in a single-step `Pipeline` named `"riesz"`, then pass a list of fully configured instances as the value for that key. `GridSearchCV` calls `pipe.set_params(riesz=candidate)`, fits, and scores each one.

```{python}
from sklearn.model_selection import cross_val_predict, GridSearchCV
from sklearn.pipeline import Pipeline
from rieszboost import RieszBooster
from riesznet import RieszNet
from forestriesz import ForestRieszRegressor

estimand = ATE()

# Five fully configured candidates across three learner families.
candidates = [
    # Boosting — depth 3, early stopping
    RieszBooster(
        estimand=estimand,
        n_estimators=2000,
        max_depth=3,
        early_stopping_rounds=20,
        validation_fraction=0.2,
        learning_rate=0.05,
    ),
    # Boosting — depth 5, early stopping
    RieszBooster(
        estimand=estimand,
        n_estimators=2000,
        max_depth=5,
        early_stopping_rounds=20,
        validation_fraction=0.2,
        learning_rate=0.05,
    ),
    # Neural net — small MLP with early stopping
    RieszNet(
        estimand=estimand,
        hidden_sizes=(64, 64),
        epochs=500,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    # Neural net — larger MLP with early stopping
    RieszNet(
        estimand=estimand,
        hidden_sizes=(128, 64),
        epochs=500,
        early_stopping_rounds=20,
        validation_fraction=0.2,
    ),
    # Forest Riesz
    ForestRieszRegressor(
        estimand=estimand,
        n_estimators=500,
        min_samples_leaf=10,
        random_state=0,
    ),
]

# Single-step Pipeline so GridSearchCV can swap in any candidate as the step.
pipe = Pipeline([("riesz", candidates[0])])

# Inner CV: 3 folds, selects the best learner+config from the five candidates.
cv_inner = GridSearchCV(
    pipe,
    param_grid=[{"riesz": [c]} for c in candidates],
    cv=3,
    n_jobs=1,
    refit=True,
)

# Outer CV: 5 cross-fitting folds. Each fold:
#   1. Fits cv_inner on 4/5 of the data (inner CV happens here, on that 4/5).
#   2. Predicts the held-out 1/5 using the winner from step 1.
# The inner model selection never touches the held-out slice.
alpha_oof = cross_val_predict(cv_inner, df, cv=5)
print(f"out-of-fold alpha shape: {alpha_oof.shape}")
print(f"first 5 values:          {alpha_oof[:5]}")
```

The call to `cross_val_predict` is identical to the single-learner case on this page. The only addition is the `Pipeline` wrapper and the list-valued `param_grid` entry — everything else is standard sklearn.

::: {.callout-note title="R users"}
The `Pipeline` + `GridSearchCV` pattern is standard Python. In R, call into Python via `reticulate` using the same code, passing an R data frame with `reticulate::r_to_py(df)` and reading back the result with `as.numeric(reticulate::py_to_r(...))`.
:::

## Other sklearn integrations

Every Riesz estimator exposes the standard `BaseEstimator` interface:

- `clone(estimator)` returns a fresh unfitted copy with the same hyperparameters.
- `Pipeline` wraps a feature-engineering step in front of the estimator.
- `HalvingGridSearchCV`, `RandomizedSearchCV`, `BayesSearchCV` (skopt) work without modification.
- `get_params()` / `set_params()` are used by the search classes to introspect and mutate the configuration.

## See also

- [`rieszboost/examples/lee_schuler/binary_dgp.py`](https://github.com/rieszreg/rieszreg/blob/main/rieszboost/examples/lee_schuler/binary_dgp.py) builds an EEE estimator using `cross_val_predict` to produce out-of-fold $\hat\alpha$.
- [Boosting](backends/boosting.qmd), [kernel](backends/kernel.qmd), [forest](backends/forest.qmd), and [neural](backends/neural.qmd) backend pages cover backend-specific tuning grids.
